---
title: "Further Changepoint Analysis"
author: "Rebecca Killick(r.killick@lancs.ac.uk)"
date: "NHS Workshop 2021"
output:
  beamer_presentation:
    includes:
      in_header: header.tex
    theme: lancaster
---

```{r,include=FALSE}
if(!require(changepoint)){
  install.packages('changepoint')
}
library(changepoint)
if(!require(EnvCpt)){
  install.packages('EnvCpt')
}
library(EnvCpt)
if(!require(changepoint.influence)){
  install.packages('changepoint.influence')
}
library(changepoint.influence)
if(!require(changepoint.geo)){
  install.packages('changepoint.geo')
}
library(changepoint.geo)
if(!require(ecp)){
  install.packages('ecp')
}
library(ecp)
knitr::opts_chunk$set(fig.align='center',fig.show='hold',fig.width=4,fig.height=4,size='footnotesize', cache=TRUE)
knitr::opts_knit$set(progress = FALSE,verbose = FALSE)
```

## Workshop Plan
* Recap of changepoints
* Checking assumptions
* Autocorrelation
* Multivariate changepoints
* Influence

There will be tasks throughout the sections.

## Recall: changepoints
For data $y_1, \ldots, y_n$, if a changepoint exists at $\tau$, then $y_1,\ldots,y_{\tau}$ differ from $y_{\tau+1},\ldots,y_n$ in some way.  

There are many different types of change.
  
```{r, echo=F, out.width='.3\\textwidth'}
par(mar=c(4,4,.3,.3)) 
set.seed(1)
# Change in mean example following EFK
x=1:500
y=c(rnorm(100,1,sd=0.5),rnorm(150,0,sd=0.5),rnorm(200,2,sd=0.5),rnorm(50,0.5,sd=0.5))
plot(x,y,type='l',xlab='',ylab='')
lines(x=1:100,y=rep(1,100),col='red',lwd=3)
lines(x=101:250,y=rep(0,150),col='red',lwd=3)
lines(x=251:450,y=rep(2,200),col='red',lwd=3)
lines(x=451:500,y=rep(0.5,50),col='red',lwd=3)
# Change in variance example following EFK
x=1:500
y=c(rnorm(100,0,sd=0.1),rnorm(150,0,sd=0.7),rnorm(200,0,sd=0.25),rnorm(50,0,sd=1))
plot(x,y,type='l',xlab='',ylab='')
# Change in regression
x=1:500
y=c(0.01*x[1:100],1.5-0.02*(x[101:250]-101),(10^-5)*(-150000+2.5*(x[251:450]^2-251^2)-(x[251:450]-250)),rep(1,50))
ynoise=y+rnorm(500,0,0.2)
plot(x,ynoise,type='l',xlab='',ylab='')
lines(x=1:100,y=0.01*x[1:100],lwd=3,col='red')
lines(x=101:250,y=1.5-0.02*(x[101:250]-101),lwd=3,col='red')
lines(x=251:450,y=(10^-5)*(-150000+2.5*(x[251:450]^2-251^2)-(x[251:450]-250)),lwd=3,col='red')
lines(x=451:500,y=rep(1,50),lwd=3,col='red')
```

## Packages
Today we will use the following packages

`library(changepoint)`

`library(EnvCpt)`

`library(changepoint.influence)`

`library(changepoint.geo)`

Other notable `R` packages are available for changepoint analysis including

* `ecp` - for univariate and multivariate energy test statistics 
* `InspectChangepoint` - for multivariate Inspect projection direction mean only change
* `hdbinseg` - for multivariate double CUSUM test statistic
* `BayesProject` - for multivariate changepoints

## Checking Assumptions
The main assumptions for a Normal likelihood ratio test for a change in mean are:

* Independent data points;
* Normal distributed points pre and post change;
* Constant variance across the data.

How can we check these?

## How to check
* Check the residuals
```{r}
set.seed(1)
m1=c(rnorm(100,0,1),rnorm(100,5,1))
m1.amoc=cpt.mean(m1)


means=param.est(m1.amoc)$mean
m1.resid=m1-rep(means,seg.len(m1.amoc))
shapiro.test(m1.resid)
```

## Residual Check
```{r}
ks.test(m1.resid,pnorm,mean=mean(m1.resid),sd=sd(m1.resid))
```

## Residual Check
```{r,out.height='0.6\\textheight'}
qqnorm(m1.resid)
qqline(m1.resid)
```

## Residual Check
```{r,out.height='0.6\\textheight'}
acf(m1.resid)
```

## Autocorrelation

What effect does autocorrelation have on our analysis?

```{r, out.height='0.6\\textheight'}
set.seed(879123)
x=c(rnorm(100),rnorm(100,3))
plot(cpt.meanvar(x,method='PELT'))
```

## Autocorrelation +ve

What effect does autocorrelation have on our analysis?

```{r, out.height='0.5\\textheight'}
source('sim.cpt.AR1.R')
set.seed(879123)
x=sim.cpt.AR1(cpts=c(0,100,200),X=cbind(rep(1,200)),init=0,
    beta=rbind(c(0,0.9),c(1,0.9)),sig2=(1-0.9^2),nsim=1)
plot(cpt.meanvar(x,method='PELT'))
```

## Autocorrelation -ve

What effect does autocorrelation have on our analysis?

```{r, out.height='0.6\\textheight'}
set.seed(879123)
x=sim.cpt.AR1(cpts=c(0,100,200),X=cbind(rep(1,200)),init=0,
    beta=rbind(c(0,-0.9),c(1,-0.9)),sig2=(1-0.9^2),nsim=1)
plot(cpt.meanvar(x,method='PELT'))
```

## Exercise
Take a look at the "Lai2005fig4" data in the `changepoint` package.
Fit changes in mean as below, then check the residuals.  

Are the assumptions of our model reasonable?

```{r}
data("Lai2005fig4")
out=cpt.mean(Lai2005fig4$GBM29,method='PELT')
```

Don't forget to look at the data!

## EnvCpt
EnvCpt automatically fits 12 different models to your data:

* Flat mean (+AR1, +AR2, +Change, +AR1+Change, +AR2+Change)
* Trend mean (+AR1, +AR2, +Change, +AR1+Change, +AR2+Change)

AR1= autoregressive of order 1 = current data point is strongly related to the last data point.

**BONUS**: Can see which model is best

**PITFALL**: Might be best to use another model which isn't checked - always look at the fit!

## EnvCpt: Example
```{r}
set.seed(879123)
x=sim.cpt.AR1(cpts=c(0,100,200),X=cbind(rep(1,200)),init=0,
    beta=rbind(c(0,0.9),c(1,0.9)),sig2=(1-0.9^2),nsim=1)
out=envcpt(x)
which.min(BIC(out))
```
## EnvCpt: Example
```{r, out.height='0.7\\textheight', out.width='\\textwidth'}
plot(out)
```

## EnvCpt: Example
```{r, out.height='0.5\\textheight'}
cpts(out$meanar1cpt)
plot(out[[which.min(BIC(out))+1]])
abline(v=cpts(out$meanar1cpt),col='blue')
```
## EnvCpt: A&E Gallstone
HES Data on monthly proportion of A&E admissions for gallstone disease from Jan 2010 - Dec 2019.
```{r,fig.height=3,fig.width=7,out.height='0.35\\textheight',out.width='\\textwidth'}
load('AnEGallstoneProp.Rdata')
plot(AnEGallstoneProp[,2],type='l')
```

Use EnvCpt to see if there is evidence for changes in the monthly proportion of A&E admissions for gallstone disease.

## Gallstone Solution
```{r}
out=envcpt(AnEGallstoneProp[,2])
which.min(BIC(out))
```

## Gallstone Solution
```{r, out.height='0.7\\textheight', out.width='\\textwidth'}
plot(out)
```
## Gallstone Solution
```{r, out.height='0.6\\textheight'}
AnEGallstoneProp$Date[cpts(out$trendar1cpt)]
plot(out$trendar1cpt)
```

## Exercise

Go back to the "Lai2005fig4" data.  Use `envcpt()` to identify the best model.  Does this fit with what we observed previously?


## Influence

* Which data points are *influential* for obtaining the segmentation? 
  * Changepoints versus Outliers
  * How to measure influence?
* How *stable* is the obtained segmentation?

## Influence: Example
```{r, out.height='0.7\\textheight'}
set.seed(30) 
x=c(rnorm(50),rnorm(50,mean=5),rnorm(1,mean=15),
  rnorm(49,mean=5),rnorm(50,mean=4))
xcpt=cpt.mean(x,method='PELT')
plot(xcpt,cpt.width=3,ylab='')
```
## How to measure?
Sources of Inspiration:

* Regression Analysis: Measures of Influence (e.g., Cook's distance)
* Robust Statistics: Influence Functions

Two routes:

* Modifying an observation
* Leaving out an observation

## Modify
\includegraphics[scale=0.45]{changemeanoutlier} 

## Stability Dashboard: Out
```{r, out.height='0.7\\textheight'}
x.inf.out=influence(xcpt,method='outlier')
out.Stability=StabilityOverview(x,cpts(xcpt),x.inf.out,
  legend.args=list(display=TRUE,x="topright",y=NULL,cex=1,
  horiz=FALSE,xpd=FALSE,bty='n'))
```

## Location Stability: Out
```{r, out.height='0.7\\textheight'}
out.location=LocationStability(cpts(xcpt),x.inf.out,
  type='Difference')
```

## Parameter Stability: Out
```{r, out.height='0.7\\textheight'}
ParameterStability(x.inf.out,original.mean=rep(
  param.est(xcpt)$mean,times=diff(c(0,xcpt@cpts))))
```

## Influence Map: Out

```{r}
out.map=InfluenceMap(cpts(xcpt),x.inf.out)
```

## Stability Dashboard: Del
```{r, out.height='0.7\\textheight'}
x.inf.del=influence(xcpt,method='delete')
del.Stability=StabilityOverview(x,cpts(xcpt),x.inf.del,
  legend.args=list(display=TRUE,x="topright",y=NULL,cex=1,
  horiz=FALSE,xpd=FALSE,bty='n'))
```

## Location Stability: Del
```{r, out.height='0.7\\textheight'}
del.location=LocationStability(cpts(xcpt),x.inf.del,
  type='Difference')
```

## Parameter Stability: Del
```{r, out.height='0.7\\textheight'}
ParameterStability(x.inf.del,original.mean=rep(
  param.est(xcpt)$mean,times=diff(c(0,xcpt@cpts))))
```

## Influence Map: Del

```{r}
del.map=InfluenceMap(cpts(xcpt),x.inf.del)
```


## Multivariate changes

In moving to the multivariate setting a number of different scenarios could arise.

* The process in each channel could be unconnected to the rest (i.e.
repeated use of univariate cpt methods might be appropriate);
* There may be some shared structure across channels. For example
  * Changes occur at the same time in all channels;
  * Changes occur in a subset of channels at the same time.
* The nature of the change could vary from one channel to another;
* \ldots and doubtless many more scenarios!


## Multivariate changes
In the multivariate setting we encounter new challenges:

* Computational expense.
* Sparsity of changepoints.
* Incorporating multivariate power.

\includegraphics[height=0.43\textheight]{multivariatecolour}

## Many methods

Some well known multivariate changepoint approaches include:

* [ecp:](https://www.jstatsoft.org/article/view/v062i07) James, Matteson (2015)
* [Inspect:](https://rss.onlinelibrary.wiley.com/doi/full/10.1111/rssb.12243) Wang, Samworth (2017)
* [DoubleCUSUM:](https://projecteuclid.org/euclid.ejs/1468849969) Cho (2016)


## GeomCP Intuition

\includegraphics[height=0.8\textheight]{GeometricIntuition}

## GeomCP Mean Ex

```{r, out.height='0.6\\textheight'}
set.seed(1)
Y=rbind(matrix(rnorm(100*20),ncol=20),cbind(matrix(rnorm(100*10),
    ncol=10),matrix(rnorm(100*10,1),ncol=10)))
image(Y)
```
## GeomCP: Mean Ex

```{r, out.height='0.6\\textheight'}
res <- geomcp(Y)
plot(res)
```

## GeomCP MeanVar Ex

```{r, out.height='0.6\\textheight'}
set.seed(1)
Y=rbind(matrix(rnorm(100*20),ncol=20),cbind(matrix(rnorm(100*10,0,2),
    ncol=10),matrix(rnorm(100*10,1,2),ncol=10)))
image(Y)
```
## GeomCP: MeanVar Ex

```{r,, out.height='0.6\\textheight'}
res <- geomcp(Y)
plot(res)
```

## Task: Genetics
Analyse the ACGH Bladder Tumor data from the ecp package.  It is 2215x43 with 43 patients. How many changes do you find?

```{r, out.height='0.5\\textheight'}
data(ACGH)
image(ACGH$data)
```

## Solution: Genetics
```{r, out.height='0.6\\textheight'}
outACGH <- geomcp(ACGH$data)
plot(outACGH)
```


## Solution: Genetics
Can also take the distance and angle vectors and analyse using CROPS
```{r, results='hide',fig.keep='all',out.height='0.6\\textheight'}
outACGH.dist=cpt.meanvar(distance(outACGH),method='PELT',
    penalty='CROPS',pen.value=c(5,500))
plot(outACGH.dist,diagnostic=TRUE)
```


## Summary

* Multivariate  is interesting but still lots of challenges in the univariate space
* Lots of interesting research in the changepoint space
* Always looking for interesting problems to work on
* Reach out if you want help / guidance

## References
[PELT:](http://www.tandfonline.com/doi/abs/10.1080/01621459.2012.737745) Killick, Fearnhead, Eckley (2012)  

[EnvCpt:](https://doi.org/10.1175/JCLI-D-17-0863.1) Beaulieu, Killick (2018) 

[geomCP:](https://link.springer.com/article/10.1007/s11222-020-09940-y) Grundy, Killick (2020)

[Influence:](https://www.tandfonline.com/doi/full/10.1080/10618600.2021.2000873) Wilms, Killick, Matteson (2021+)

